Hazardous asteroids forecast via Markov random fields

### 2021-12-11 Presentation/Presentation.tex
PUB

\mode<presentation>
{
  \usetheme{Darmstadt}      % or try Darmstadt, Madrid, Warsaw, ...
  \usecolortheme{beaver} % or try albatross, beaver, crane, ...
  \usefonttheme{default}  % or try serif, structurebold, ...
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
}

\usepackage[backend=bibtex]{biblatex}
\addbibresource{bibliography.bib}
\newtheorem*{remark}{Remark}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Pic/Intro.jpg}}
\title[Hazardous asteroids forecast via Markov random fields]{Hazardous asteroids forecast via Markov random fields}
\subtitle{Project for the exam: Probabilistic Modelling (DSE)}
\author{Marzio De Corato}
\date{\today}

@@

\begin{frame}
\vspace{+4 cm}  \titlepage
\end{frame}

@@

% Uncomment these lines for an automatically generated outline.
%\begin{frame}{Outline}
%  \tableofcontents
%\end{frame}

@@

@@

@@

@@

\section{Prob. mod.}

\begin{frame}{GLASSO}
\begin{columns}
\column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Pic/GLASSO_0.1.pdf}
    \subcaption{$\rho$=0.2 \begin{center}
    \end{center}}
  \end{figure} 
  \column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Pic/GLASSO_0.2.pdf}
    \subcaption{$\rho$=0.2 \begin{center}
    \end{center}}
  \end{figure}
\end{columns}
\begin{center}
Performed with the GLASSO package \cite{glasso}
\end{center}
\end{frame}

@@

\section{Other ML alg.}
\begin{frame}{Random Forest}
\begin{columns}
\column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{Pic/RF_confusion.pdf}
  \end{figure} 
  \column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{Pic/ROC_RF.pdf}
  \end{figure}
\end{columns}
\begin{center}
Performed with the rfor package \cite{rfor}
\end{center}
\end{frame}

@@

@@

\begin{frame}{Conclusions: forecast performances vs intepretability}
\begin{itemize}
\item The mgm algorithm is not the best one in term of performances, but it provides the connections between the features. On the other side, except for the variable importance in RF, the other are black box one
\item The mgm model, as the other graphical model is open to a true scientific validation, the other not. 
\item The probabilistic models lack in the forecast is definitely compensated by their interetability 
\item This is meaningful since this two features are in conflict 
\item The probabilistic models provide a good trade-off between intepretability and forecast performances, as long as one is interest to produce a really scientific result (e.g if the only aim is the forecast the RF is definitely better. However how long one can trust to the RF result ?)
\end{itemize}
\end{frame}

\section{SI}

@@

@@

@@

\begin{frame}{mmod}
\begin{equation}
\begin{split}
f(i,y)=& p(i)(2\pi)^{-q/2}det(\Sigma)^{-1/2} \\
& exp\left[-\dfrac{1}{2}\left(y-\mu(i)\right)^{T}\Sigma^{-1}\left(y-\mu(i)\right)\right]
\end{split}
\label{gaussMix}
\end{equation}
\begin{equation}
\begin{split}
f(i,y) & = \exp\left\lbrace g(i)+\sum_{u}h^{u}(i)y_{u}-\dfrac{1}{2}\sum_{uv} y_{u}y_{v}k_{uv}\right\rbrace \\
&= \exp\left\lbrace g(i)+h(i)^{T}y-\dfrac{1}{2}y^{T}Ky \right\rbrace
\end{split}
\end{equation}

\begin{center}
where $g(i)$, $h(i)$ and $K$ are the canonical parameters
\end{center}

\end{frame}
\begin{frame}{mmod}
\begin{equation}
\begin{split}
K=&\Sigma^{-1} \\
h(i)=&\Sigma^{-1}\mu(i) \\
g(i)=&\log p(i) -\frac{1}{2}\log det (\Sigma) \\
&-\dfrac{1}{2}\mu(i)^{T}\Sigma^{-1}\mu(i)-\dfrac{q}{2}\log 2\pi
\end{split}
\end{equation}
\end{frame}

\begin{frame}{Graphical models}
\begin{theorem}[Hammersley-Clifford]
A positive distribution p(\textbf{y})>0 satisfies the CI properties of an indirect graph G iif p can be represented as a product of factor, one per maximal clique,  i.e.
\begin{equation}
p(\textbf{y}|\theta)= \dfrac{1}{Z(\theta)}\prod_{c \in C }\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
where C is the set of all the (maximal) cliques of G,  and Z($\theta$) is the partition function given by 
\begin{equation}
Z(\theta):= \sum_{y}\prod_{c\in C}\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
Note that this partition function is what ensures the overall distribution sums to 1
\end{theorem}
\end{frame}
\begin{frame}{Graphical models}
\begin{equation}
p(y|\theta)=\dfrac{1}{Z(\theta)} exp\left(-\sum_{c}E(y_{c}|\theta_{c})\right)
\end{equation}
\begin{equation}
\psi_{c}(y_{c}|\theta_{c})=exp\left(-E(y_{c}|\theta_{c})\right)
\end{equation}
\end{frame}

>>>>

\mode<presentation>
{
  \usetheme{Madrid}      % or try Darmstadt, Madrid, Warsaw, ...
  \usecolortheme{beaver} % or try albatross, beaver, crane, ...
  \usefonttheme{default}  % or try serif, structurebold, ...
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
}

\usepackage[bookmarks]{hyperref}
\usepackage[backend=bibtex]{biblatex}
\addbibresource{bibliography.bib}
\newtheorem*{remark}{Remark}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Pic/Intro.jpg}}
\title[Probabilistic modeling project]{Hazardous asteroids forecast via Markov random fields}
\subtitle{Project for the exam: Probabilistic Modelling (DSE)}
\author{Marzio De Corato}
\date{\today}

@@

\begin{frame}
\vspace{+4.2 cm}  \titlepage
\end{frame}

@@

% Uncomment these lines for an automatically generated outline.
\begin{frame}{Outline}
\setcounter{tocdepth}{1}
\begin{center}
  \tableofcontents
\end{center}
\end{frame}

@@

\begin{frame}{}
\begin{center}
{\Huge Celestial mechanics}
\end{center}
\end{frame}

@@

\begin{frame}{}
\begin{center}
{\Huge Dataset}
\end{center}
\end{frame}

@@

\begin{frame}{}
\begin{center}
{\Huge Prelim. analysis}
\end{center}
\end{frame}

@@

\section{Probabilistic modelling}

\begin{frame}{}
\begin{center}
{\Huge Probabilistic modelling}
\end{center}
\end{frame}

\begin{frame}{GLASSO}
\begin{columns}
\column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Pic/GLASSO_0.1.pdf}
    \subcaption{$\rho$=0.1 \begin{center}
    \end{center}}
  \end{figure} 
  \column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Pic/GLASSO_0.2.pdf}
    \subcaption{$\rho$=0.2 \begin{center}
    \end{center}}
  \end{figure}
\end{columns}
\begin{center}
Performed with the GLASSO package \cite{glasso}
\end{center}
\end{frame}

@@

\section{Other ML algorithms}

\begin{frame}{}
\begin{center}
{\Huge Other ML algorithms}
\end{center}
\end{frame}

\begin{frame}{Random Forest}
\begin{columns}
\column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{Pic/RF_confusion.pdf}
  \end{figure} 
  \column{0.5\textwidth}
  \begin{figure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{Pic/ROC_RF.pdf}
  \end{figure}
\end{columns}
\begin{center}
Performed with the rfor package \cite{rfor}
\end{center}
\end{frame}

@@

\begin{frame}{}
\begin{center}
{\Huge Conclusions}
\end{center}
\end{frame}

@@

\begin{frame}{Conclusions: forecast performances vs intepretability}
\begin{itemize}
\item The mgm algorithm is not the best one in term of performances, but it provides the connections between the features. On the other side, except for the variable importance in RF, the other are black box one
\item The mgm model, as the other graphical model is open to a true scientific validation, the other not. 
\item The probabilistic models lack in the forecast is definitely compensated by their interpretability 
\item This is meaningful since this two features are in conflict 
\item The probabilistic models provide a good trade-off between intepretability and forecast performances, as long as one is interest to produce a really scientific result (e.g if the only aim is the forecast the RF is definitely better. However how long one can trust to the RF result ?)
\end{itemize}
\end{frame}

\begin{frame}

{\Large \textit{In its efforts to learn as much as possible about nature, modern physics has found that certain things can never be “known” with certainty. Much of our knowledge must always remain uncertain. The most we can know is in terms of probabilities.} Richard P. Feynman (1918-1988)}
\end{frame}

\section{Supporting Info}

\subsection{ROC and $\phi$ factor}
\begin{frame}{}
\begin{center}
{\Huge ROC and $\phi$ factor}
\end{center}
\end{frame}

@@

\subsection{FAMD}
\begin{frame}{}
\begin{center}
{\Huge FAMD}
\end{center}
\end{frame}

@@

\subsection{Graphical models}
\begin{frame}{}
\begin{center}
{\Huge Graphical models}
\end{center}
\end{frame}

\begin{frame}{Graphical models}
\begin{theorem}[Hammersley-Clifford]
A positive distribution p(\textbf{y})>0 satisfies the CI properties of an indirect graph G iif p can be represented as a product of factor, one per maximal clique,  i.e.
\begin{equation}
p(\textbf{y}|\theta)= \dfrac{1}{Z(\theta)}\prod_{c \in C }\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
where C is the set of all the (maximal) cliques of G,  and Z($\theta$) is the partition function given by 
\begin{equation}
Z(\theta):= \sum_{y}\prod_{c\in C}\psi_{c}(\textbf{y}_{c}|\theta_{c})
\end{equation}
Note that this partition function is what ensures the overall distribution sums to 1
\end{theorem}
\end{frame}

\begin{frame}{Graphical models}
\begin{equation}
p(y|\theta)=\dfrac{1}{Z(\theta)} exp\left(-\sum_{c}E(y_{c}|\theta_{c})\right)
\end{equation}
\begin{equation}
\psi_{c}(y_{c}|\theta_{c})=exp\left(-E(y_{c}|\theta_{c})\right)
\end{equation}
\end{frame}

@@

\begin{frame}{mmod - Homogeneus Mixed Interaction (HMI) models}
\begin{equation}
f(\textbf{i},\textbf{y})= p(\textbf{i})(2\pi)^{-q/2}det(\Sigma)^{-1/2}\exp\left[-\dfrac{1}{2}\left(\textbf{y}-\mu(\textbf{i})\right)^{T}\Sigma^{-1}\left(\textbf{y}-\mu(\textbf{i})\right)\right]
\label{gaussMix}
\end{equation}

\begin{equation}
f(\textbf{i},\textbf{y})=\exp\left[g(\textbf{i}) +h(\textbf{i})^{T}\textbf{y}-\frac{1}{2}y^{T}K\textbf{y}\right]
\end{equation}

\begin{center}
where $K$ is the concentration matrix, $g(i)$ and $h(i)$ are the log-linear expansion of the probability $p(\textbf{i})$ (canonical parameters)
\end{center}

\end{frame}
\begin{frame}{mmod}
\begin{center}
Decomposable model
\end{center}
\begin{equation}
\hat{f}(x)=\prod_{j=1}^{k}\dfrac{\hat{f}_{C_{j}}(x_{C}_{j})}{\hat{f}_{S_{j}}(x_{S}_{j})} 
\end{equation}
Likehood 
\begin{equation}
\hat{\mathcal{L}_{s}}=\sum_{i}n(i)log\left(\dfrac{n(i)}{N}\right)-Nq\dfrac{log(2pi)}{2}-N\dfrac{log(det(S))}{2}-N\frac{q}{2}
\end{equation}

\subsection{Information theory}
\begin{frame}{}
\begin{center}
{\Huge Information theory}
\end{center}
\end{frame}
